---
title: "Complex Survey Designs: Design-Based Standard Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complex Survey Designs: Design-Based Standard Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(cli.unicode = FALSE, useFancyQuotes = FALSE)
# gt tables self-contained: avoids an HTML dependency in the pandoc header
# that some Windows rmarkdown setups write in native encoding (breaks pandoc).
suppressWarnings(suppressMessages({library(knitr); library(gt)}))
try(utils::assignInNamespace("knit_print.gt_tbl", function(x, ...)
  knitr::asis_output(gsub("\u00b2", "2", gsub("\u00a0", " ", as.character(gt::as_raw_html(x))))),
  ns = "gt"), silent = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("survey", quietly = TRUE)
)
```

## From weighted point estimates to defensible ones

A weighted mean is not enough for a paper: you need a standard error that
respects the sampling design (strata, clusters, finite-population correction).
`rquasar` builds this on top of the
[`survey`](https://cran.r-project.org/package=survey) package — so you get the
correct numbers in one call, and they match `survey` **exactly**. `rquasar`
does not reinvent Lumley's estimators; it removes the boilerplate around them.

This vignette uses the classic `api` data (California school Academic
Performance Index) shipped with `survey`.

```{r data}
library(rquasar)
data(api, package = "survey")   # loads apistrat, apiclus1, apisrs, ...

# apistrat: a stratified sample with sampling weights (pw), strata (stype),
# and finite-population correction (fpc)
str(apistrat[, c("api00", "meals", "ell", "pw", "stype", "fpc")])
```

## 1. Declare the design once

```{r design}
qsr_config()
qsr_data(apistrat)
qsr_validate(checks = "survey")

# Stratified by school type, weighted by pw, with the FPC:
qsr_svydesign(strata = "stype", weights = "pw", fpc = "fpc")
```

## 2. Design-based tabulation: estimate, SE, CI, deff

Ask `qsr_tabulate()` for `se = TRUE` and it returns the mean API by school type
with a design-based standard error, a 95% confidence interval and the design
effect — Taylor linearization, via the design just declared:

```{r tab}
qsr_tabulate(by = "stype", measure = "api00", stat = "mean", se = TRUE, digits = 1)
```

These are the same numbers `survey::svyby()` produces — to machine precision:

```{r check}
des <- survey::svydesign(ids = ~1, strata = ~stype, weights = ~pw,
                         fpc = ~fpc, data = apistrat)
as.data.frame(survey::svyby(~api00, ~stype, des, survey::svymean, vartype = "se"))
```

## 3. Design-based regression

Pass `design = TRUE` to fit through `survey::svyglm()`, so the coefficients'
standard errors account for the design (not just the weights):

```{r model}
qsr_model(api00 ~ meals + ell, design = TRUE)
qsr_table(caption = "Table 1. Design-based regression (svyglm)")
```

For a replicate-weight design (BRR / jackknife / bootstrap), build one with
`survey::svrepdesign()` and pass it straight to `design =`. The default
`qsr_tabulate()` (without `se`) stays a fast, dependency-free point estimate;
`se = TRUE` is the opt-in that makes it publication-grade.
