---
title: "Complex Survey Designs: Validate, Model, and Hand Off"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complex Survey Designs: Validate, Model, and Hand Off}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(cli.unicode = FALSE, useFancyQuotes = FALSE)
# gt tables self-contained: avoids an HTML dependency in the pandoc header
# that some Windows rmarkdown setups write in native encoding (breaks pandoc).
suppressWarnings(suppressMessages({library(knitr); library(gt)}))
try(utils::assignInNamespace("knit_print.gt_tbl", function(x, ...)
  knitr::asis_output(gsub("\u00b2", "2", gsub("\u00a0", " ", as.character(gt::as_raw_html(x))))),
  ns = "gt"), silent = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("survey", quietly = TRUE)
)
```

## Know the division of labor

`rquasar` speeds up the *plumbing* around survey microdata — validation,
harmonization, and publication-ready output. It is **not** a design-based
inference engine, and it does not pretend to be. For estimates that account for
weights, strata, and finite-population corrections, the
[`survey`](https://cran.r-project.org/package=survey) package is the standard,
and `rquasar` hands off to it.

This vignette uses the classic `api` data (California school Academic
Performance Index) shipped with `survey`.

```{r data}
library(rquasar)
data(api, package = "survey")   # loads apistrat, apiclus1, apisrs, ...

# apistrat: a stratified sample with sampling weights (pw), strata (stype),
# and finite-population correction (fpc)
str(apistrat[, c("api00", "meals", "ell", "pw", "stype", "fpc")])
```

## 1. Validate with rquasar

```{r validate}
qsr_config()
qsr_data(apistrat)
qsr_validate(checks = "survey")
```

## 2. A quick descriptive model (unweighted)

`qsr_model()` fits an ordinary model — useful for exploration, but it ignores
the sampling design:

```{r model}
qsr_model(api00 ~ meals + ell)
qsr_table(caption = "Table 1. OLS (unweighted, exploratory)")
```

## 3. Hand off to `survey` for design-based estimates

When the numbers have to be correct for the design, build a design object and
estimate with `survey`. `rquasar` does not do this for you — it points you here:

```{r svy}
design <- survey::svydesign(
  ids = ~1, strata = ~stype, weights = ~pw, fpc = ~fpc, data = apistrat
)
survey::svymean(~api00, design)
summary(survey::svyglm(api00 ~ meals + ell, design = design))$coefficients
```

The difference between the two model tables is the sampling design at work —
which is exactly why design-based inference belongs in `survey`, not in a
convenience wrapper.
