---
title: "Machine Learning with QUASAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine Learning with QUASAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

QUASAR provides a unified interface for 12 machine learning methods through
`qsr_ml()`, cross-validation through `qsr_cv()`, and model comparison through
`qsr_compare()`. Every method works with a single function call and integrates
with the QUASAR context engine.

## Supported Methods

| Method | Engine | Type |
|--------|--------|------|
| `"rf"` | ranger | Random Forest |
| `"xgboost"` | xgboost | Gradient Boosting |
| `"svm"` | e1071 | Support Vector Machine |
| `"nn"` | nnet | Neural Network |
| `"knn"` | class | K-Nearest Neighbors |
| `"lasso"` | glmnet | Lasso (L1) |
| `"ridge"` | glmnet | Ridge (L2) |
| `"elastic"` | glmnet | Elastic Net |
| `"tree"` | rpart | Decision Tree |
| `"nb"` | e1071 | Naive Bayes |
| `"kmeans"` | stats | K-Means Clustering |
| `"pca"` | stats | Principal Components |

## Quick Start

```{r quickstart}
library(rquasar)

qsr_config(random_seed = 42)
qsr_data(mtcars)

# Fit a Random Forest
qsr_ml(mpg ~ ., method = "rf")

# Cross-validate
qsr_cv(mpg ~ ., method = "rf", k = 5)
```

## Comparing Multiple Methods

```{r compare}
library(rquasar)
data(wage1, package = "wooldridge")

qsr_config(random_seed = 42)
qsr_data(wage1)

# Fit statistical model
qsr_model(lwage ~ educ + exper + tenure, name = "ols")

# Fit ML models
qsr_ml(lwage ~ educ + exper + tenure, method = "rf", name = "forest")
qsr_ml(lwage ~ educ + exper + tenure, method = "nn", size = 10, name = "nnet")

# Compare all models
qsr_compare()
```

## Cross-Validation Workflow

```{r cv}
qsr_data(mtcars)

# Compare methods via CV
qsr_cv(mpg ~ wt + hp + disp, method = "rf", k = 10)
qsr_cv(mpg ~ wt + hp + disp, method = "tree", k = 10)
qsr_cv(mpg ~ wt + hp + disp, method = "nn", k = 10, size = 5)
```

`qsr_cv()` automatically detects regression vs. classification based on the
response variable. Regression reports RMSE and R-squared; classification
reports accuracy.

## Unsupervised Learning

```{r unsupervised}
qsr_data(iris[, 1:4])

# K-Means clustering
qsr_ml(method = "kmeans", centers = 3)

# PCA
qsr_ml(method = "pca")
```

## Passing Method-Specific Arguments

All extra arguments are forwarded to the underlying engine:

```{r args}
# Random Forest with 1000 trees
qsr_ml(mpg ~ ., method = "rf", num.trees = 1000)

# XGBoost with custom parameters
qsr_ml(mpg ~ ., method = "xgboost", nrounds = 200, max_depth = 6)

# Neural Network with 20 hidden units
qsr_ml(mpg ~ ., method = "nn", size = 20, maxit = 500)

# Lasso with specific lambda range
qsr_ml(mpg ~ ., method = "lasso")
```

## Integration with Publication Output

After fitting ML models, use `qsr_plot()` for visualization:

```{r output}
qsr_data(mtcars)
qsr_ml(mpg ~ ., method = "rf")

# Variable importance plot
qsr_plot(type = "importance")
```

## Design Decisions

- **No preprocessing required.** QUASAR handles formula parsing, matrix
  conversion for XGBoost/glmnet, and factor encoding internally.
- **Argument filtering.** When using Random Forest via `qsr_cv()`, arguments
  that are not recognized by ranger are silently filtered to prevent warnings.
- **Reproducibility.** Seeds are managed by the context engine. Set once with
  `qsr_config(random_seed = 42)` and all downstream models use it.
