---
title: "From Raw Data to APA7 in 10 Lines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Raw Data to APA7 in 10 Lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(cli.unicode = FALSE, useFancyQuotes = FALSE)
# gt tables self-contained: avoids an HTML dependency in the pandoc header
# that some Windows rmarkdown setups write in native encoding (breaks pandoc).
suppressWarnings(suppressMessages({library(knitr); library(gt)}))
try(utils::assignInNamespace("knit_print.gt_tbl", function(x, ...)
  knitr::asis_output(gsub("\u00b2", "2", gsub("\u00a0", " ", as.character(gt::as_raw_html(x))))),
  ns = "gt"), silent = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 5
)
```

## The Problem

Every data analysis project in R starts the same way: 200 lines of
boilerplate. Set a seed. Configure paths. Load data. Fit a model. Format a
table for publication. Style a plot. Export to Word. Repeat next project.

**QUASAR eliminates that cycle.** One framework for
data -> insights -> output.

## Quick Start: A Complete Analysis

```{r quickstart}
library(rquasar)

# 1. Configure once
qsr_config()

# 2. Register data
qsr_data(mtcars)

# 3. Fit a model (one line -- data comes from context)
qsr_model(mpg ~ wt + hp + disp)

# 4. Publication-ready table (zero arguments -- reads model from context)
qsr_table()

# 5. Coefficient plot (zero arguments -- reads model from context)
qsr_plot(type = "coefficient")
```

That's it. Five function calls. No `broom::tidy()`, no `gt::tab_options()`,
no 15-line ggplot2 theme blocks. QUASAR handles it.

## The Four Layers

QUASAR is organized in four layers, each solving a specific pain point:

| Layer | What it does | Key functions |
|-------|-------------|---------------|
| **Scaffold** | Standardized project structure | `qsr_init()` |
| **Context** | Configure once, use everywhere | `qsr_config()`, `qsr_data()`, `qsr_model()` |
| **Connectors** | Databases, APIs, Spark, Python | `qsr_db()`, `qsr_fetch()`, `qsr_spark()`, `qsr_python()` |
| **Output** | Tables, plots, reports | `qsr_table()`, `qsr_plot()`, `qsr_report()` |

## Layer 1: Scaffold

Create a standardized project in one call:

```{r scaffold, eval = FALSE}
qsr_init(
  name   = "electoral_analysis_2024",
  type   = "academic",
  author = "Jane Doe"
)
```

This generates a ready-to-use directory structure with numbered folders
(`01_data/`, `02_transform/`, `03_model/`, `04_output/`), YAML config files,
and a `main.R` entry point.

## Layer 2: Context -- The Heart of QUASAR

The context is what makes zero-argument calls possible. Register your data
and configuration once -- every downstream function reads from it
automatically.

```{r context}
# Configure analysis parameters
qsr_config(
  significance_level = 0.05,
  random_seed        = 42,
  output_format      = "apa7"
)

# Register data -- all functions now know about it
qsr_data(mtcars)

# Fit models -- stored in context by name
qsr_model(mpg ~ wt + hp, name = "base")
qsr_model(mpg ~ wt + hp + disp + qsec, name = "full")

# Everything downstream works with zero arguments
qsr_table()                       # table from last model
qsr_table(type = "summary")      # descriptive stats from active data
qsr_plot(type = "coefficient")    # coefficient plot from last model
qsr_plot(type = "correlation")    # correlation matrix from active data
```

## Layer 3: Connectors

### Databases

```{r db, eval = FALSE}
# SQLite -- zero setup
qsr_db("sqlite", dbname = "research.sqlite", table = "surveys")

# PostgreSQL -- credentials from environment variables
qsr_db("postgres", dbname = "research",
       query = "SELECT * FROM surveys WHERE year > 2020")
```

After `qsr_db()`, the data is automatically registered in context. You can
immediately call `qsr_model()` or `qsr_table(type = "summary")`.

### Academic APIs

```{r fetch, eval = FALSE}
# World Bank GDP per capita
qsr_fetch("worldbank", indicator = "NY.GDP.PCAP.CD",
          country = "all", start = 2000, end = 2023)

# FRED unemployment rate
qsr_fetch("fred", series_id = "UNRATE")

# Any CSV from a URL
qsr_fetch("url", url = "https://example.com/data.csv")
```

Results are cached for 24 hours to avoid redundant API calls during
iterative analysis.

### Python Interop

```{r python, eval = FALSE}
# Call any Python function
qsr_python("math", fn = "sqrt", args = list(16))

# Fit a scikit-learn model
qsr_python("sklearn.linear_model", fn = "LinearRegression",
           fit = list(X_train, y_train), predict = X_test)
```

## Layer 4: Output Engine

### Tables

`qsr_table()` creates publication-ready tables with APA7 or Chicago
formatting:

```{r tables_demo}
library(rquasar)

model <- lm(mpg ~ wt + hp, data = mtcars)
qsr_table(model, caption = "Table 1. OLS Estimates of Fuel Efficiency")
```

Export to multiple formats simultaneously:

```{r tables_export, eval = FALSE}
qsr_table(model, export = c("html", "latex", "word"))
```

### Plots -- Static and Interactive

QUASAR offers 20 plot types through a single function:

```{r plot_scatter}
qsr_plot(mtcars, type = "scatter", x = wt, y = mpg, color = cyl,
         title = "Weight vs. Fuel Efficiency")
```

```{r plot_coef}
model <- lm(mpg ~ wt + hp + disp + qsec, data = mtcars)
qsr_plot(model, type = "coefficient",
         title = "Figure 1. OLS Coefficient Estimates")
```

Make any plot interactive with one flag:

```{r interactive, eval = FALSE}
qsr_plot(mtcars, type = "scatter", x = wt, y = mpg,
         interactive = TRUE)
```

Native plotly types for advanced visualizations:

```{r plotly_native, eval = FALSE}
# 3D scatter
qsr_plot(mtcars, type = "scatter3d", x = wt, y = hp, z = mpg, color = cyl)

# Correlation heatmap (interactive)
qsr_plot(mtcars, type = "heatmap")

# Radar chart
qsr_plot(team_stats, type = "radar", color = team)
```

### Three Themes

```{r themes, eval = FALSE}
qsr_plot(mtcars, type = "scatter", x = wt, y = mpg,
         theme = "quasar_academic")   # serif, clean, publication-ready
qsr_plot(mtcars, type = "scatter", x = wt, y = mpg,
         theme = "quasar_minimal")    # no grid, axis lines only
qsr_plot(mtcars, type = "scatter", x = wt, y = mpg,
         theme = "quasar_dark")       # dark background for presentations
```

## The Complete Pipeline

Here is a realistic academic workflow -- from database to publication -- in
10 lines:

```{r pipeline, eval = FALSE}
library(rquasar)

qsr_config(output_format = "apa7")
qsr_db("sqlite", dbname = "study_data.sqlite", table = "responses")
qsr_model(satisfaction ~ age + income + education)
qsr_table(caption = "Table 1. OLS Estimates")
qsr_table(type = "summary", caption = "Table 2. Descriptive Statistics")
qsr_plot(type = "coefficient", title = "Figure 1. Effect Sizes")
qsr_plot(type = "correlation", title = "Figure 2. Correlation Matrix",
         interactive = TRUE, export = "html")
qsr_report(title = "Customer Satisfaction Analysis", journal = "r_journal")
```

Ten lines. Database to publication-ready report. No boilerplate.

## Rust-Powered Performance

QUASAR includes a Polars/Rust backend that streams operations over a file
instead of loading it into R. Its edge is **peak memory** (out-of-core work),
not raw speed -- for data that fits in RAM, `data.table` is faster, and QUASAR
says so. See the *Fast Data with the Rust Backend* vignette and the
[benchmarks](https://github.com/TodoEconometria/rquasar-benchmarks) for measured,
honest numbers. The user never sees Rust -- the API stays the same.

```{r rust_read, eval = FALSE}
# Read a 50M-row CSV -- 10-30x faster than read.csv()
qsr_read("data/large_file.csv")

# Fast summary stats without loading the full file into R
qsr_fast_summary("data/large_file.csv")

# Group-by aggregation directly on the file
qsr_fast_group("data/sales.csv", by = "region", col = "revenue", fn = "mean")

# Filter 100M rows in milliseconds
qsr_fast_filter("data/logs.csv", col = "status", op = "eq", value = 200)
```

### Adaptive Process Isolation

For files larger than 128MB, QUASAR automatically runs Polars in a
separate subprocess using all available CPU cores. This eliminates
memory conflicts between Rust and R while maximizing parallelism.
Smaller files use a direct single-threaded call with zero overhead.

```{r benchmark, eval = FALSE}
# Compare QUASAR vs base R on your own data
qsr_benchmark("data/big_file.csv")

# Reports wall-clock and peak memory. See the benchmarks repo for the honest
# comparison against data.table and dplyr.
```

## Design Philosophy

- **Convention over configuration** -- sensible defaults, override when needed
- **Zero-argument by default** -- if context knows it, you don't type it
- **Compatible, not competitive** -- works with tidyverse, does not replace it
- **Invisible performance** -- Rust backend under the hood, same clean API

## Learn More

- GitHub: [TodoEconometria/R-QUASAR](https://github.com/TodoEconometria/R-QUASAR)
- Author: Juan Marcelo Gutierrez Miranda (TodoEconometria)
