---
title: "Fast Data with the Rust Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast Data with the Rust Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(cli.unicode = FALSE, useFancyQuotes = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("nycflights13", quietly = TRUE)
)
```

## Operations that never fully load into R

`rquasar` ships a Rust backend (extendr + Polars) that runs aggregations,
filters and column selection **directly on a file**, streaming it rather than
reading it whole into memory. This vignette uses `nycflights13::flights`
(336,776 rows) written to a temporary CSV, which stands in for the multi-gigabyte
files the backend is built for.

```{r write}
library(rquasar)

csv <- file.path(tempdir(), "flights.csv")
write.csv(nycflights13::flights, csv, row.names = FALSE)
```

## Group-by aggregation on the file

Mean departure delay per carrier — computed by Polars on the file, only the
result returns to R:

```{r group}
qsr_fast_group(csv, by = "carrier", col = "dep_delay", fn = "mean")
```

## Filter on the file

Keep only long-haul flights (distance > 2500 miles) without materializing the
rest:

```{r filter}
long_haul <- qsr_fast_filter(csv, col = "distance", op = "gt", value = 2500)
nrow(long_haul)
```

## Read a fast preview

Read the first rows, keeping only the columns you ask for:

```{r read}
qsr_read(csv, columns = c("carrier", "dep_delay", "distance"), n_rows = 5)
```

## Where it wins, and where it does not

The backend's real advantage is **peak memory**: because the data never fully
lands in R, it uses a small fraction of the memory of `data.table` or `dplyr` —
the edge that matters when a file does not fit comfortably in RAM. On
**wall-clock**, `data.table` is faster on data that already fits in memory, and
`rquasar` does not claim otherwise. Measured, reproducible timings (wall-clock
**and** peak memory) against `data.table` and `dplyr` live in the companion
repository,
[rquasar-benchmarks](https://github.com/TodoEconometria/rquasar-benchmarks).
