---
title: "Setup: Installation, Requirements, and AI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup: Installation, Requirements, and AI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Requirements

| Component | Needed for | Version |
|---|---|---|
| **R** | everything | >= 4.2 |
| **Rust toolchain** (`cargo`, `rustc`) + `xz` | the Polars/Rust backend (`qsr_read`, `qsr_fast_*`) | rustc >= 1.65 |
| `pdftools` / `haven` | reading PDF / SPSS-Stata files in the AI helpers | Suggests |

Install the Rust toolchain with [rustup](https://rustup.rs). On **Windows** also
install Rtools and use the GNU toolchain
(`rustup default stable-x86_64-pc-windows-gnu`); on macOS/Linux the default host
toolchain works.

## Installation

```{r install}
# development version
# install.packages("pak")
pak::pak("TodoEconometria/R-QUASAR/quasar")

# once on CRAN
install.packages("rquasar")
```

## Verify it works

```{r verify}
library(rquasar)

# context engine
qsr_config()
qsr_data(mtcars)
qsr_model(mpg ~ wt + hp)
qsr_table()

# Rust backend (operates on a file)
tmp <- tempfile(fileext = ".csv")
write.csv(mtcars, tmp, row.names = FALSE)
qsr_fast_group(tmp, by = "cyl", col = "mpg", fn = "mean")
```

## AI-assisted workflows

`rquasar` includes optional AI helpers for survey work:

- `qsr_ai_review()` — compare two codebooks / questionnaires across waves.
- `qsr_ai_classify()` — classify free-text fields (crops, ethnicities) into
  standard categories, handling typos and dialect.
- `qsr_ai_flag()` — surface data-quality issues from summary statistics.

They are entirely optional: the rest of the package works without them.

### Two backends

**1. Anthropic (Claude) — the default.** Set your key once per session (or in
`.Renviron`):

```{r anthropic}
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")

qsr_ai_classify(
  data       = survey_data,
  column     = "crop_name",
  categories = c("coca", "coffee", "cacao", "banana", "other")
)
```

**2. Local (Ollama) — runs on your machine, no API key, nothing leaves your
computer.** Install [Ollama](https://ollama.com) and pull a model
(`ollama pull llama3.1`), then:

```{r ollama}
qsr_ai_classify(
  data       = survey_data,
  column     = "crop_name",
  categories = c("coca", "coffee", "cacao", "banana", "other"),
  provider   = "ollama"
)

# pick the model and host if you like:
Sys.setenv(QSR_OLLAMA_MODEL = "llama3.1",
           OLLAMA_HOST = "http://localhost:11434")
```

### Privacy

- **`provider = "anthropic"` sends your data to Anthropic's API.** Do **not** use
  it on confidential or personal microdata.
- **`provider = "ollama"` keeps everything local** — no data is transmitted and
  no key is used. Prefer it for sensitive data.
- The Anthropic key is read from the environment and sent only as a request
  header; it is never printed, logged, or returned in any result object.
