---
title: "From Survey Microdata to Publication Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Survey Microdata to Publication Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The Applied Economist's Problem

Working with household survey microdata (ENAHO, Encuesta de Hogares, GEIH)
involves weeks of boilerplate: harmonizing variable names across waves,
cleaning free-text fields, converting currencies to constant USD, validating
data quality, and formatting regression tables for journal submission.

QUASAR was built from direct experience with these datasets. Every function
in the survey module solves a real problem encountered while processing
1.14 million observations across Bolivia and Peru household surveys.

## Complete Workflow

```{r workflow}
library(rquasar)

qsr_config(
  significance_level = 0.05,
  random_seed = 42,
  output_format = "apa7"
)

# 1. Load survey data
qsr_data(survey_panel)

# 2. Clean text fields (fuzzy matching for typos)
qsr_normalize_text(
  column = "crop_name",
  dictionary = c("coca", "cafe", "cacao", "banana", "arroz"),
  max_dist = 2
)

# 3. Harmonize geographic codes across waves
qsr_crosswalk(
  column = "ubigeo",
  crosswalk = "data/ubigeo_crosswalk_2017.csv",
  year_column = "year",
  apply_before = 2017
)

# 4. Convert to constant USD
qsr_currency(
  columns = c("ingreso_laboral", "ingreso_total"),
  from_currency = "PEN",
  base_year = 2015
)

# 5. Validate
qsr_validate(checks = "survey")

# 6. Estimate
qsr_model(log_income ~ education + experience + indigenous, type = "lm")

# 7. Publish
qsr_table(caption = "Table 1. OLS Estimates of Income Determinants")
```

## Text Normalization

Survey respondents write crop names, occupations, and locations with typos,
abbreviations, and dialectal variants. `qsr_normalize_text()` uses
Levenshtein distance to standardize them:

```{r normalize}
qsr_normalize_text(
  column = "occupation",
  dictionary = c("agricultor", "comerciante", "minero",
                 "profesor", "obrero", "empleado"),
  max_dist = 2,
  review = TRUE  # prints matches for manual verification
)
```

## Geographic Crosswalks

Administrative boundaries change between census years (Peru 2017, Colombia
DIVIPOLA updates). `qsr_crosswalk()` maps old codes to new ones:

```{r crosswalk}
qsr_crosswalk(
  column = "cod_municipio",
  crosswalk = data.frame(
    old = c("05001", "05002"),
    new = c("05001", "05899")
  )
)
```

## Currency Conversion

Built-in exchange rates and CPI deflators for BOB, PEN, COP, and BRL
(2006-2023, source: World Bank):

```{r currency}
qsr_currency(
  columns = c("ingreso_laboral"),
  from_currency = "BOB",
  base_year = 2015
)
```

Custom rates can be passed for any other currency or when official rates
differ from market rates.

## Data Validation

Three presets designed for survey microdata:

```{r validate}
# Basic: no empty dataset, no all-NA columns, no duplicates
qsr_validate(checks = "basic")

# Survey: adds weight checks, NA rate, negative monetary values
qsr_validate(checks = "survey")

# Panel: adds year column, balance across waves
qsr_validate(checks = "panel")

# Custom checks
qsr_validate(checks = list(
  "No duplicate IDs" = function(df) {
    if (any(duplicated(df$id))) "Duplicates" else TRUE
  }
))
```

## Decision Logging

Every non-obvious data cleaning decision should be documented. QUASAR
generates structured markdown logs for replication packages:

```{r decision_log}
qsr_decision_log(
  id = "BOL-D001",
  title = "Income reference period correction",
  variable = "ingreso_laboral",
  waves = "2012 onwards",
  problem = "Reference period changed from monthly to weekly",
  evidence = "Mean income 2011=1847 BOB vs 2012=412 BOB",
  decision = "Multiply 2012+ incomes by 4.3",
  code_ref = "harmonize_bol.R line 87",
  impact_n = "None (no observations dropped)",
  alternative = "Exclude 2012-2023 — rejected: loses 11 years"
)
```

## Publication Tables

After estimation, generate journal-ready tables:

```{r tables}
# Single model table
qsr_model(lwage ~ educ + exper + tenure)
qsr_table(caption = "Table 1. Mincer Equation")

# Compare multiple specifications
qsr_model(lwage ~ educ, name = "bivariate")
qsr_model(lwage ~ educ + exper, name = "controls")
qsr_model(lwage ~ educ + exper + tenure, name = "full")
qsr_compare()

# Export to LaTeX for journal submission
qsr_table(export = "latex")
```

## Download Guides

For surveys requiring manual download (INE Bolivia, INEI Peru, DANE
Colombia), QUASAR provides step-by-step guidance:

```{r guide}
qsr_guide("ine_bolivia", year = 2023)
qsr_guide("inei_peru")
qsr_guide("dane_colombia")
```
