---
title: "From Survey Microdata to Publication Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Survey Microdata to Publication Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The applied economist's problem

Working with household survey microdata (ENAHO, Encuesta de Hogares, GEIH)
involves weeks of boilerplate: harmonizing variable names across waves, cleaning
free-text fields, converting currencies to constant USD, validating data
quality, and formatting regression tables for journal submission. Every function
in the survey module solves a real problem encountered while processing
1.14 million observations across Bolivia and Peru household surveys.

## A familiar example first

The pipeline is the same on any microdata. Here it is on `wage1` from the
`wooldridge` package — the dataset behind the textbook Mincer wage equation —
so the workflow is recognizable before we reach the survey-specific tools:

```{r wage1, eval = requireNamespace("wooldridge", quietly = TRUE)}
library(rquasar)
data("wage1", package = "wooldridge")

qsr_config()
qsr_data(wage1)
qsr_model(lwage ~ educ + exper + tenure)
qsr_table(caption = "Table 1. Mincer wage equation (wooldridge::wage1)")
```

The rest of this vignette uses `qsr_survey` to demonstrate the tools built
specifically for developing-country household survey microdata.

## A dataset you can run

This vignette runs on `qsr_survey`, a **fully synthetic** household-survey panel
shipped with the package. It mirrors the public structure of ENAHO/EH — `ubigeo`
geographic codes, income in local currency, expansion weights, indigenous
self-identification and free-text fields with realistic data-entry noise — but
contains **no real observation**. See `?qsr_survey`.

```{r data}
library(rquasar)

qsr_config(significance_level = 0.05, random_seed = 42)
qsr_data(qsr_survey)

str(qsr_survey[, c("year", "ubigeo", "crop_name", "ingreso_laboral",
                   "education", "indigenous")])
```

## 1. Clean text fields (fuzzy matching)

Respondents write crop names with typos and dialectal variants.
`qsr_normalize_text()` uses edit distance to standardize them into a new
`crop_name_clean` column (the original is preserved for audit):

```{r normalize}
d <- qsr_normalize_text(
  column = "crop_name",
  dictionary = c("coca", "cafe", "cacao", "banana", "arroz"),
  max_dist = 2,
  review = FALSE
)

# dozens of noisy variants collapse to 5 clean categories
table(d$crop_name_clean)
```

## 2. Harmonize geographic codes across waves

Administrative boundaries change between census years. `qsr_crosswalk()` maps old
codes to new ones, here only for waves before the 2017 update, into
`ubigeo_harmonized`:

```{r crosswalk}
d <- qsr_crosswalk(
  column = "ubigeo",
  crosswalk = data.frame(
    old = c("0502", "0602"),
    new = c("0503", "0699")
  ),
  year_column = "year",
  apply_before = 2017
)

table(pre_2017 = d$year < 2017,
      remapped = d$ubigeo_harmonized %in% c("0503", "0699"))
```

## 3. Convert to constant USD

Built-in exchange rates and CPI deflators for BOB, PEN, COP and BRL (2006-2023,
source: World Bank). Converted columns get a `_usd` suffix:

```{r currency}
d <- qsr_currency(
  columns = c("ingreso_laboral", "ingreso_total"),
  from_currency = "PEN",
  base_year = 2015
)

head(d[, c("ingreso_laboral", "ingreso_laboral_usd")])
```

## 4. Validate

Three presets designed for survey microdata. The `"survey"` preset adds weight
checks, NA-rate and negative-monetary checks — `qsr_survey` carries a few coding
errors on purpose, so the report flags them:

```{r validate}
qsr_validate(checks = "survey")
```

## 5. Estimate and publish

The model reads from context; `qsr_table()` produces an APA7 regression table.
The synthetic data is generated from a Mincer process, so the estimates are
economically sensible — a positive return to schooling and a negative ethnic
income gap:

```{r model}
qsr_model(log_income ~ education + experience + indigenous)
```

```{r table}
qsr_table(caption = "Table 1. OLS Estimates of Income Determinants")
```

## Comparing specifications

```{r compare}
qsr_model(log_income ~ education, name = "bivariate")
qsr_model(log_income ~ education + experience, name = "controls")
qsr_model(log_income ~ education + experience + indigenous, name = "full")
qsr_compare()
```

## Decision logging

Every non-obvious cleaning decision should be documented for replication.
`qsr_decision_log()` generates structured markdown; point it at a temporary
directory when running non-interactively:

```{r decision_log, eval = FALSE}
qsr_decision_log(
  id = "SURV-D001",
  title = "Income reference period correction",
  variable = "ingreso_laboral",
  waves = "2012 onwards",
  problem = "Reference period changed from monthly to weekly",
  evidence = "Mean income drops sharply between adjacent waves",
  decision = "Rescale post-2012 incomes to a monthly basis",
  code_ref = "make_qsr_survey.R",
  impact_n = "None (no observations dropped)",
  alternative = "Drop the affected waves - rejected: loses coverage"
)
```

## Download guides

For surveys requiring manual download (INE Bolivia, INEI Peru, DANE Colombia),
`qsr_guide()` prints step-by-step instructions:

```{r guide, eval = FALSE}
qsr_guide("ine_bolivia", year = 2023)
qsr_guide("inei_peru")
```
